---
title: "Reproducible research_project1"
output: html_document
---
## Introdution
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Load and Process Data
Include required libraries
```{r echo = TRUE, results='hide'}
library(dplyr)
library(lubridate)
library(ggplot2)
```

Unzip the downloaded activity.zip file and read the file. The data can also be downloaded from [url](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

```{r echo = TRUE}
unzip("activity.zip")
act_data <- read.table("activity.csv",sep = ",",header = TRUE)
```

The class of date column needs to be corrected to Date.
```{r echo = TRUE}
act_data$date <- ymd(act_data$date)
```

## Plotting and Reporting data

### Calculate total number of steps taken each day and show it using histogram
```{r echo = TRUE}
res1_sum <- summarise(group_by(act_data,date),steps = sum(steps))
res1_hist <- ggplot(res1_sum,aes(x=date,y=steps))
res1_hist + geom_histogram(stat="identity")
```

Calculate mean and median for total number of steps taken per day
```{r echo = TRUE}
res1_mean <- mean(res1_sum$steps,na.rm = TRUE)
print(paste("The mean per day is:",res1_mean))
res1_median<- median(res1_sum$steps,na.rm = TRUE)
print(paste("The median per day is:",res1_median))
```

### A time series plot (i.e.type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo = TRUE}
res2_mean <- summarise(group_by(act_data,interval),steps = mean(steps,na.rm = TRUE))
res2_ts <- ggplot(res2_mean,aes(x=interval,y=steps))
res2_ts + geom_line()
```

The 5-minute interval, which on average across all the days in the dataset, contains the maximum number of steps is obtained using following code.
```{r echo = TRUE}
res2_max <- summarise(group_by(act_data,interval),steps = max(steps,na.rm = TRUE))
maxsteps <- res2_max[res2_max$steps == max(res2_max$steps),]
print(paste("The interval containing max steps:",maxsteps$interval))
```

### Impute missing values
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NA's)
```{r echo = TRUE}
tot_miss_val <- sum(!complete.cases(act_data))
print(paste("Total missing values are:",tot_miss_val))
```

Devise a strategy for filling in all of the missing values in the dataset. Impute missing values with the average of steps taken on the given day
```{r echo = TRUE}
act_data_noNA <- act_data
res3_mean <- summarise(group_by(act_data,date),steps = mean(steps))

for (i in 1:nrow(act_data_noNA)){
    if (is.na(act_data_noNA$steps[i])){
        res3 <- filter(res3_mean,date == act_data_noNA$date[i])
        act_data_noNA$steps[i] <- res3$steps
        
        if (is.na(act_data_noNA$steps[i])){
          act_data_noNA$steps[i] <- 0
        
        }
    }
}
```

plot histogram of the total number of steps taken each day. This is after imputing missing values.
```{r echo = TRUE}
res4_sum <- summarise(group_by(act_data_noNA,date),steps = sum(steps))
res4_hist <- ggplot(res4_sum,aes(x=date,y=steps))
res4_hist + geom_histogram(stat="identity")
```

Calculate and report the mean and median total number of steps taken per day after imputing missing values.
```{r echo = TRUE}
res4_mean <- mean(res4_sum$steps)
print(paste("The mean per day is:",res4_mean))
res4_median <- median(res4_sum$steps)
print(paste("The median per day is:",res4_mean))
```

The difference in mean and median of data, before and after imputing missing values shows teh variation

### New factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r echo = TRUE}
res5_data <- act_data
res5_data$day <- weekdays(act_data$date)

res5_data$day <- gsub("Monday|Tuesday|Wednesday|Thursday|Friday","Weekday",res5_data$day)
res5_data$day <- gsub("Saturday|Sunday","Weekend",res5_data$day)
res5_data$day <- as.factor(res5_data$day)
```

### Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r echo = TRUE}
res5_mean <- aggregate(steps ~ interval + day, res5_data, mean)
res6_mean <- summarise(group_by(res5_data,interval,day),steps = mean(steps,na.rm = TRUE))
res5_ts <- ggplot(res5_mean,aes(x=interval,y=steps))
res5_ts + geom_line() +facet_grid(.~day)
```
